Orientation Control System: Enhancing Aerial Maneuvers for Quadruped Robots

### 2023-01-09 answersToReviewers.tex

%\noindent This is an obligatory section in “Advances in Respiratory Medicine”, whose goal is to increase the discoverability and readability of the article via search engines and other scholars. Highlights should not be a copy of the abstract, but a simple text allowing the reader to quickly and simplified find out what the article is about and what can be cited from it. Each of these parts should be devoted up to 2~bullet points.\vspace{3pt}\\
%\textbf{What are the main findings?}
% \begin{itemize}[labelsep=2.5mm,topsep=-3pt]
% \item First bullet.
% \item Second bullet.
% \end{itemize}\vspace{3pt}
%\textbf{What is the implication of the main finding?}
% \begin{itemize}[labelsep=2.5mm,topsep=-3pt]
% \item First bullet.
% \item Second bullet.
% \end{itemize}
%}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Dear Ms. Vera Popovic,\\
Thank you for giving us the chance to submit a revised draft version of the manuscript titled "Orientation Control System: Enhancing Aerial Maneuvers for Quadruped Robots" to the MDPI journal \textit{Sensors}. We are grateful for the time and the effort you and the reviewers have dedicated to providing insightful comments on the original manuscript. We have incorporated changes to reflect most of the provided suggestions: they are highlighted in red within the manuscript. For removing the highlights, please comment line 73 and uncomment line 74 in the file named \texttt{mdpi2022roscia.tex}.\\
Here is a point-by-point response to the reviewers' comments and concerns.

\section*{Comments for Reviewer 1}
\begin{itemize}
	\item \textbf{Comment 1:} \textit{Extend the text of the manuscript (e.g. introduction or conclusion) with specific results in the world and Europe, - Improve the quality of the paper by presenting the results of publications by researchers and experts who are involved in this field and registered in world databases (wos). These are e.g: Measurement of industrial robot pose repeatability, Investigation of snake robot locomotion possibilities in a pipe, thanks.}
	\item[] \textbf{Response:} 
	\MF{Thank you for the constructive comment. We have extended the introduction of the manuscript with additional explanations and references to papers about the specific research of orientation control of legged robots during aerial maneuvers. The authors of the additionally cited papers are world experts and the majority of the newly added papers are listed in the Web of Science. The most recent publications of 2022 are not yet listed there. We analysed the two suggested examples of potentially relevant publications to add as a reference. We concluded that they are not directly related to the contributions of our work, because we tackle the angular dynamics of the main body of a quadruped robot, that is observable.}	\\
	\item \textbf{Comment 2:} \textit{Figure 4 should be contrasting and readable}
	\item[] \textbf{Response:} Thank you for pointing this out. We have incorporated your suggestion in the revised manuscript. \\
	\item \textbf{Comment 3:} \textit{ Conclusions and future work should be extended to contain practical applications based on research described in this paper - expand references.}
	\item[] \textbf{Response:} We are glad you made such an important hint to improve the quality of the manuscript. 
	\MF{In the conclusions section we have added the following paragraph to discuss the possible practical applications.}\\	
	Possible applications of the presented OCS include but are not limited to efficiently adjust the posture of quadruped robots walking or jumping on uneven terrains. As proved in the third simulation (back-flip), our approach improves the capabilities of quadrupeds in space applications enabling fast locomotion by means of leaps, ensuring a reactive control action on the robot angular momentum. Furthermore, the method presented in Section 3 for designing the OCS does not depend on the specific platform, thus it can be replicated for reorienting mechanical structures with different morphology, e.g, monopods or bipeds. A construction worker's backpack can contain two flywheels with incident rotation axes: in the event of a fall from scaffolding, they can be used to reorient the human body to avoid the impact of the head with the ground with the same controller proposed in this work. \\
	\MF{We also expanded the number of references, please refer to Comment 5 below.}\\
	\item \textbf{Comment 4:} \textit{Number all mathematical equations}
	\item[] \textbf{Response:} Thank you for bringing this to our attention. We applied the comment into the document.\\	
	\item \textbf{Comment 5:} \textit{For article type, 20 references are not enough. Please add more references (>20) during your revisions.}
	\item[] \textbf{Response:} We really appreciate your suggestion, which helped us to improve the manuscript. We included other relevant works in the literature review, e.g. [An, 2022], [Tang, 2022], and [Kurtz, 2022].\\
	\item \textbf{Comment 6:} \textit{Standardization of literature input (into one style)}
	\item[] \textbf{Response:} Agree. We have corrected the literature input.\\
\end{itemize}

### 2023-01-18 mdpi2022roscia_v7/sensors-2114505.tex
for final review

%=================================================================
\documentclass[sensors,article,submit,pdftex,moreauthors]{Definitions/mdpi} 
% For posting an early version of this manuscript as a preprint, you may use "preprints" as the journal and change "submit" to "accept". The document class line would be, e.g., \documentclass[preprints,article,accept,moreauthors,pdftex]{mdpi}. This is especially recommended for submission to arXiv, where line numbers should be removed before posting. For preprints.org, the editorial staff will make this change immediately prior to posting.

@@

%=================================================================
% MDPI internal commands
\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother
\pubvolume{1}
\issuenum{1}
\articlenumber{0}
\pubyear{2022}
\copyrightyear{2022}
%\externaleditor{Academic Editor: Firstname Lastname}
\datereceived{} 
%\daterevised{} % Only for the journal Acoustics
\dateaccepted{} 
\datepublished{} 
%\datecorrected{} % Corrected papers include a "Corrected: XXX" date in the original paper.
%\dateretracted{} % Corrected papers include a "Retracted: XXX" date in the original paper.
\hreflink{https://doi.org/} % If needed use \linebreak
%\doinum{}
%------------------------------------------------------------------
% The following line should be uncommented if the LaTeX file is uploaded to arXiv.org
%\pdfoutput=1

@@

\newcommand{\MF}[1]{\textcolor{red}{#1}}
%\newcommand{\MF}[1]{\textcolor{black}{#1}}
\newcommand{\CS}[1]{\textcolor{violet}{\textbf{csemini}: #1}}
\newcommand{\FR}[1]{\textcolor{teal}{\textbf{froscia}: #1}}
\newcommand{\ADP}[1]{\textcolor{blue}{\textbf{adelprete}: #1}}
%=================================================================
%% Please use the following mathematics environments: Theorem, Lemma, Corollary, Proposition, Characterization, Property, Problem, Example, ExamplesandDefinitions, Hypothesis, Remark, Definition, Notation, Assumption
%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).

@@

% Authors, for the paper (add full first names)
\Author{Francesco Roscia $^{1}$, Andrea Cumerlotti $^{1, \, 2}$, Andrea Del Prete $^{2}$, Claudio Semini $^{1}$ and Michele Focchi$^{1, \, 3, ^*}$}

@@

% MDPI internal command: Authors, for citation in the left column
\AuthorCitation{Roscia, F.; Cumerlotti, A.; Del Prete, A., Semini, C.; Focchi, M.}
% If this is a Chicago style journal: Lastname, Firstname, Firstname Lastname, and Firstname Lastname.

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Dynamic Legged Systems (DLS) lab, Istituto Italiano di Tecnologia (IIT), Genoa, Italy\\
$^{2}$ \quad Industrial Engineering Department (DII), University of Trento, Trento, Italy\\
$^{3}$ \quad Department of Information Engineering and Computer Science (DISI), University of Trento, Trento, Italy\\
}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{For legged robots, aerial motions are the only option to overpass obstacles that cannot be circumvent with standard locomotion gaits. In these cases, the robot must perform a leap to either jump onto the obstacle or fly over it. However, these movements represent a challenge because during the flight phase the \gls{com} cannot be controlled, and the robot orientation has limited controllability. This paper focuses on the latter issue and proposes an \gls{ocs} consisting of two rotating and actuated masses (flywheels or reaction wheels) to gain control authority on the robot orientation. Because of the conservation of angular momentum, their rotational velocity can be adjusted to steer the robot orientation even when there are no contacts with the ground. The axes of rotation of the flywheels are designed to be incident, leading to a compact orientation control system that is capable of controlling both roll and pitch angles, considering the different moment of inertia in the two directions. \MF{The concept is tested with simulations on the robot Solo12.}}

\makeglossaries
% Keywords
\keyword{Legged robot, Orientation Control, Articulated Multi-body System}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}\label{sec:introduction}
Legged robots are designed for traversing rough terrain.
Different types of gaits, such as trot \MF{\cite{barasuol2013reactive}} or crawl \MF{\cite{focchi2020heuristic}}, have been developed for quadrupedal robots. 
Thanks to the progress of the last two decades, robots have become lighter and stronger, which has enabled them to perform agile locomotion.
However, sometimes it is not possible to get around or over an obstacle with the gaits mentioned above, and jumps are required.

When the robot is in the air, the \gls{com} moves on the ballistic trajectory, that is completely defined by the lift-off position and velocity. On the other hand, the base orientation can be changed exploiting the conservation of the system angular momentum. This means that it is possible to control the base angular velocity by changing the inertia of the robot, e.g., changing the joints configuration. Nevertheless, the majority of quadrupeds are designed with light legs, resulting in limbs that have small influence on the total angular momentum.

Quadrupedal animals, like cats, can rearrange tail and trunk to correct the orientation during a fall \cite{kane1969dynamical}.
% Tail
Many works in the field of robotics used an additional link as a tail, like in \cite{chu2019null} and \cite{wenger2016frontal}.
This link rotates around an axis that does not pass through the robot \gls{com}: the distances between the axis of rotation and the \gls{com} of both trunk and tail result in a large effect on the total inertia even with a small tail mass.
However, the placement of the additional link makes the resulting robot asymmetric. Moreover, due to its limited range of motion, a tail can be used only for a single jump, not for a repeated sequence \cite{johnson2012tail}.
\MF{To circumvent these drawbacks, in \cite{an2022design}, \cite{an2020development}, \cite{tang2022towards} the authors attach a morphable inertial tail with 3-\gls{dofs} (pitch, yaw, and telescoping) on a monopod, a biped and a quadruped, respectively, to enhance more agile locomotion and improve safe landing.}
\\
% Legs reconfig
It is possible to obtain a similar result by creating repetitive circular motions with the feet, like in \cite{hoffman2021exploiting}. \MF{Moving the feet outwards increases the robot total inertia, so if a leg is extended during half of the motion and retracted in the other half, a net angular momentum will result on the trunk, causing a rotation. In \cite{kurtz2022mini}} the authors proposed special heavy boots for Mini Cheetah and used a neural network to calculate online joint trajectories. However, this solution unnecessarily increases the inertia of the legs, complicating the locomotion problem, which can no longer rely on the massless leg assumption. \\ 
% Gyros
Another option is to use a \gls{cmg}.
It consists of a wheel, spinning at a constant angular velocity inside two or three actuated gimbals.
Tilting the wheel's axis of rotation generates the gyroscopic torque.
This system is widespread in spacecrafts reorientation \cite{yoon2002spacecraft}, but less frequently exploited in robot locomotion, either wheeled \cite{brown1996single} or legged \cite{mikhalkov2021gyrubot}.
The \gls{cmg} presents interesting capabilities, but the presence of a pan-tilt unit to drive the gyroscope makes it impractical to mount it on a small, lightweight robot.\\
% flywheel
Flywheels represent an additional option for controlling the robot orientation.
Changing the angular velocity of a rotating mass attached to the trunk generates a torque that can reorient the robot. This device has been first applied for spacecraft orientation \cite{oland2009reaction}, and later it has been only sporadically investigated in legged locomotion for controlling pitch orientation, both for bipeds \cite{Brown2016}, \cite{xiong2020sequential} and quadrupeds \cite{kolvenbach2019towards}, \cite{vasilopoulos2016quadruped}.
Compared with tails, flywheels do not have position limits, and since their rotation axes pass through the \gls{com}, its angular momentum results holonomic \cite{machairas2015quadruped}.
To get a fast response, it is necessary to have an abrupt change in the flywheels angular velocity (angular acceleration).
Using a brake avoids the employment of a motor able to deliver higher torques \cite{gajamohan2012cubli}, 
keeping the system compact: the motor slowly accelerates the wheel to a certain speed to store angular momentum and, when a reorientation is required, the break stops its spin.
Since the effect of the break is unidirectional, a limitation of this approach is given by the fact that it is possible to generate a rotation of the base only in the opposite direction of the flywheel angular velocity making them unsuitable for applications where \textit{continuous} controllability is required. On the other hand, direct-drive controlled flywheels enable to create accelerations in both directions,  and  to implement continuous control laws. \\
The benefits of this kind of \gls{ocs} are multiple. 
\begin{enumerate}
\item Flywheels can correct orientation errors during the flight (e.g. due to disturbances or tracking inaccuracies in the angular momentum achieved at the lift-off)   % and non-idealities).  
in a continuous manner. 
\item They enable to track a time-varying reference, e.g. to have the robot land with a desired angular velocity (possibly zero) and orientation.
\item  Even in presence of contacts, they can enhance the landing phase or stabilize dynamic gaits (e.g. trot), by significantly reducing trunk oscillations.  
\item  The presence of these additional joints, whose only function is to control the orientation, gives the possibility to relieve the effort of the legs.
\end{enumerate}
In more complex scenarios, like for a somersault, limbs and \gls{ocs} can operate in parallel to achieve a rotation angle larger than the one achievable only with legs (e.g., due to torque limitation).

\subsection{Proposed Approach and Contribution}
\begin{figure}[t!]
	\centering
	\includegraphics[width=7cm,height=8cm,keepaspectratio]{figures/solo_fw_pic_compressed.png}
	\caption{\small The proposed Orientation Control System for 
		for the $2.5 \ \mathrm{kg}$ quadruped robot Solo12 consists of two $0.1 \ \mathrm{kg}$ flywheels with incident rotation axes. Each wheel is located in a 3D-printed shell and mounted on the trunk body.}
	\label{fig:solo12flywheels}
\end{figure} 
\MF{This work introduces} a compact \gls{ocs} based on two flywheels, mounted on the trunk of the lightweight quadruped Solo12 \cite{grimminger2020open}. \MF{The open-source hardware and software are reported in \cite{github_solo}. Fig. \ref{fig:solo12flywheels} shows the \gls{ocs} mounted on the trunk of the robot. }
The contributions of the paper are:
\begin{itemize}
	\item the design of a novel \gls{ocs} that enables to control the orientation of a legged robot during a jump in an effective way, while keeping the design simple. 
	In particular, the axes of rotation of the flywheels are set to be incident, enabling \textit{continuous} controllability in both the directions (roll and pitch) while keeping the device compact.
	\item simulations  with the quadruped Solo12 that demonstrate the effectiveness of the proposed approach.
\end{itemize}

\subsection{Outline}
The remainder of this paper is organized as follow. In Section \ref{sec:background}, the law of conservation of the total angular momentum is recalled. In Section \ref{sec:ocs}, the design principles are presented together with the strategy for simultaneously control the robot roll and pitch orientations. Section \ref{sec:simulations} presents simulation in different scenarios that demonstrate the the capability of our \gls{ocs} to reject disturbances and to track angular references when there is no contact with the ground, and damping of base oscillation after the touch-down. Conclusions and a possible evolution of the work ar reported in Section \ref{sec:conclusions}.

\section{Background}
\label{sec:background}
The starting point for any \gls{ocs} is the Euler's equation. For any mechanical system, the time derivative of the angular momentum $\bm{L}$ computed with the respect to a reference point $O$, fixed in an inertial frame, equals the sum of the moments $\bm{M}_i$ applied to the system with respect to the same reference point:
\begin{equation}
\dot{\bm{L}} = \sum_i \bm{M}_i
\end{equation}
When the resultant of the external moments applied to the system is zero, the Euler's equation simplifies to:
\begin{equation}
\dot{\bm{L}} = 0 \quad \Rightarrow \quad \bm{L}(t) = const,
\label{eq:cons_ang_mom}
\end{equation}
which is known as conservation of angular momentum.
Referring to legged robots, this condition occurs when the system is not in contact with the ground or other objects, e.g., during a fall or the flight phase of a jump.
In this case, it is possible to change the angular velocity of the base link changing the joint positions and velocities, as a result of the \textit{non-holonomy} of the total angular momentum \cite{wieber2016modeling}: if the angular momentum of a certain body changes, the one of the others must change to maintain the total sum constant.
\subsection{Preliminary Analysis}
Most of quadruped robots are designed in such a way that the largest amount of the mass is located at the main body. As a consequence, the contribution of the moments due to the acceleration of other bodies (i.e. leg links) is moderate. \\
As an example, let consider the case of Solo12. Even if each limb accounts for about $13.4\%$ of the total mass, changes of the total robot inertia come with the motion of only the upperleg (thigh) and the lowerleg (calf), since the displacement of distribution of the hip mass with the system \acrshort{com} is approximately constant. \MF{In view of this, each limb is responsible for only $7.5\%$ of the total mass to vary the total inertia of the robot}. Flywheels can be used for alleviating the lack of control authority.

\section{Orientation Control System}
\label{sec:ocs}
In this section, a procedure to select the inertia of the flywheels \MF{is presented, together with a strategy} to exploit them for simultaneously controlling the robot roll and pitch orientation.

>>>>

%=================================================================
\documentclass[sensors,article,accept,pdftex,moreauthors]{Definitions/mdpi} 
% For posting an early version of this manuscript as a preprint, you may use "preprints" as the journal and change "submit" to "accept". The document class line would be, e.g., \documentclass[preprints,article,accept,moreauthors,pdftex]{mdpi}. This is especially recommended for submission to arXiv, where line numbers should be removed before posting. For preprints.org, the editorial staff will make this change immediately prior to posting.

@@

%=================================================================
% MDPI internal commands
\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother
\pubvolume{1}
\issuenum{1}
\articlenumber{0}
\pubyear{2023}
\copyrightyear{2023}
%学术编辑+dates
\externaleditor{Academic Editor: Emanuele Lindo Secco 
}
\datereceived{6 December 2022 %MDPI: Please add the date.
}
\daterevised{6 January 2023 %MDPI: Please add the date.
}
%\daterevised{} % Only for the journal Acoustics
\dateaccepted{16 January 2023 %MDPI: Please add the date.
}
\datepublished{{~} 
}
%\datecorrected{} % Corrected papers include a "Corrected: XXX" date in the original paper.
%\dateretracted{} % Corrected papers include a "Retracted: XXX" date in the original paper.

\hreflink{https://doi.org/} % If needed use \linebreak
%\doinum{}
%------------------------------------------------------------------
% The following line should be uncommented if the LaTeX file is uploaded to arXiv.org
%\pdfoutput=1

@@

\newcommand{\MF}[1]{\textcolor{black}{#1}}
%\newcommand{\MF}[1]{\textcolor{black}{#1}}
\newcommand{\CS}[1]{\textcolor{violet}{\textbf{csemini}: #1}}
\newcommand{\FR}[1]{\textcolor{teal}{\textbf{froscia}: #1}}
\newcommand{\ADP}[1]{\textcolor{blue}{\textbf{adelprete}: #1}}
%=================================================================
%% Please use the following mathematics environments: Theorem, Lemma, Corollary, Proposition, Characterization, Property, Problem, Example, ExamplesandDefinitions, Hypothesis, Remark, Definition, Notation, Assumption
%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).

@@

% Authors, for the paper (add full first names)
\Author{\hl{Francesco} %MDPI: Please carefully check the accuracy of names and affiliations.
 Roscia $^{1}$, Andrea Cumerlotti $^{1,2}$, Andrea Del Prete $^{2}$, Claudio Semini $^{1}$ and Michele Focchi $^{1,3,}$*}

@@

% MDPI internal command: Authors, for citation in the left column
\AuthorCitation{Roscia, F.; Cumerlotti, A.; Del Prete, A.; Semini, C.; Focchi, M.}
% If this is a Chicago style journal: Lastname, Firstname, Firstname Lastname, and Firstname Lastname.

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Dynamic Legged Systems (DLS) Lab, Istituto Italiano di Tecnologia (IIT), \hl{Genoa,} %MDPI: Please add the postal code (or ZIP code in the U.S.). If the postal code is not available, please provide the P.O. Box. The following highlights are the same.
 Italy; \linebreak  \hl{francesco.roscia@iit.it} %MDPI: We added all the emails in affiliations according to the submitting system. Please confirm.
 (F.R.); andrea.cumerlotti@studenti.unitn.it (A.C.); claudio.semini@iit.it (C.S.)\\
$^{2}$ \quad Industrial Engineering Department (DII), University of Trento, \hl{Trento}, Italy; andrea.delprete@unitn.it (A.D.P.)\\
$^{3}$ \quad Department of Information Engineering and Computer Science (DISI), University of Trento, \hl{Trento}, Italy\\
}

@@

% Abstract (Do not insert blank lines, i.e., \\) 
\abstract{For legged robots, aerial motions are the only option to overpass obstacles that cannot be circumvented with standard locomotion gaits. In these cases, the robot must perform a leap to either jump onto the obstacle or fly over it. However, these movements represent a challenge, because, during the flight phase, the Center of Mass (CoM) cannot be controlled, and there is limited controllability over the orientation of the robot. This paper focuses on the latter issue and proposes an  Orientation Control System (OCS), consisting of two rotating and actuated masses (flywheels or reaction wheels), to gain control authority on the orientation of the robot. Due to the conservation of angular momentum, the rotational velocity if the robot can be adjusted to steer the robot's orientation, even when the robot has no contact with the ground. The axes of rotation of the flywheels are designed to be incident, leading to a compact orientation control system that is capable of controlling both roll and pitch angles, considering the different moments of inertia in the two directions. \MF{The concept was tested by means of simulations on the \mbox{robot Solo12}.}}

\makeglossaries
% Keywords
\keyword{legged robot; orientation control; articulated multi-body system}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}\label{sec:introduction}
Legged robots are designed for traversing rough terrain.
Different types of gaits, such as trot \MF{\cite{barasuol2013reactive}} or crawl \MF{\cite{focchi2020heuristic}}, have been developed for quadrupedal robots. 
Thanks to progress over the last two decades, robots have become lighter and stronger, which has enabled them to perform with agile locomotion.
However, sometimes it is not possible for the robot to move around or over an obstacle with the gaits mentioned above, and jumps are required.

When the robot is in the air, the CoM moves on a ballistic trajectory, and this is completely defined by the lift-off position and velocity. On the other hand, the base orientation can be changed so as to exploit the conservation of the system's angular momentum. This means that it is possible to control the base angular velocity by changing the inertia of the robot, e.g., changing the configuration of the joints. Nevertheless, the majority of quadrupeds are designed with light legs, resulting in limbs that have little influence on the total angular momentum.

Quadrupedal animals, like cats, can rearrange their tails and trunks to correct their orientations during a fall \cite{kane1969dynamical}.
% Tail
Much work in the field of robotics has made use of an additional link, such as a tail, as in \cite{chu2019null,wenger2016frontal}.
This link rotates around an axis that does not pass through the robot's CoM. The distances between the axis of rotation and the CoM of both trunk and tail have a large effect on the total inertia, even with a small tail mass.
However, the placement of the additional link makes the resulting robot asymmetric. Moreover, due to its limited range of motion, a tail can be used only for a single jump, not for a repeated sequence \cite{johnson2012tail}.
\MF{To circumvent these drawbacks, in \cite{an2022design,an2020development,tang2022towards} the authors attached a morphable inertial tail with 3 Degrees of Freedom (DoFs) (pitch, yaw, and telescoping) on a monopod, a biped and a quadruped, respectively, to enhance the agility of locomotion and to improve safety in landing.}

% Legs reconfig
It is possible to obtain a similar result by creating repetitive circular motions with the feet, like in \cite{hoffman2021exploiting}. \MF{Moving the feet outwards increases the robot's total inertia, so if a leg is extended during half of the motion and retracted in the other half, a net angular momentum results on the trunk, causing a rotation. In \cite{kurtz2022mini}} the authors proposed special heavy boots for Mini Cheetah and used a neural network to calculate online joint trajectories. However, this solution unnecessarily increased the inertia of the legs, complicating the locomotion problem, which could no longer rely on the massless leg assumption.

% Gyros
Another option is to use a  Control Moment Gyroscope (CMG).
This consists of a wheel, spinning at a constant angular velocity, inside two or three actuated gimbals.
Tilting the wheel's axis of rotation generates gyroscopic torque.
This system is widespread in spacecraft reorientation \cite{yoon2002spacecraft}, but less frequently exploited in robot locomotion, either wheeled \cite{brown1996single} or legged \cite{mikhalkov2021gyrubot}.
The CMG presents interesting capabilities, but the presence of a pan-tilt unit to drive the gyroscope makes it impractical to mount it on a small, lig\mbox{htweight rob}ot.

% flywheel
Flywheels represent an additional option for controlling a robot's orientation.
Changing the angular velocity of a rotating mass attached to the trunk generates a torque that can reorient the robot. This device was first applied for spacecraft orientation \cite{oland2009reaction}, and only sporadically investigated in legged locomotion for controlling pitch orientation, both for bipeds \cite{Brown2016,xiong2020sequential} and quadrupeds \cite{kolvenbach2019towards,vasilopoulos2016quadruped}.
Compared with tails, flywheels do not have position limits, and, since their rotational axes pass through the CoM, angular momentum results in holonomic \cite{machairas2015quadruped}.
To get a fast response, it is necessary to have an abrupt change in the flywheel's angular velocity (angular acceleration).
Using a brake avoids the employment of a motor able to deliver higher torques \cite{gajamohan2012cubli}, 
keeping the system compact. The motor slowly accelerates the wheel to a certain speed to store angular momentum and, when a reorientation is required, the break stops its spin.
Since the effect of the break is unidirectional, a limitation of this approach lies in the fact that it is only possible to generate a rotation of the base in the opposite direction of the flywheel's angular velocity, making the approach unsuitable for applications where \textit{continuous} controllability is required. On the other hand, direct-drive controlled flywheels can create accelerations in both directions,  and can implement continuous control laws.

The benefits of this kind of OCS are multiple. 
\begin{enumerate}
\item Flywheels can correct orientation errors (e.g., due to disturbances or tracking inaccuracies in the angular momentum achieved at the lift-off) during flight    % and non-idealities).  
in a conti\mbox{nuous ma}nner. 
\item They can track a time-varying reference, e.g., the robot can land with a desired angular velocity (possibly zero) and orientation.
\item  Even in the presence of contact, the landing phase can be enhanced or dynamic gaits, such as a trot, stabilized, by significantly reducing trunk oscillations.  
\item  The presence of these additional joints, the only functions of which  are to control orientation, provides the possibility of relieving the efforts of the legs.
\end{enumerate}

In more complex scenarios, like for a somersault, limbs and OCS can operate in parallel to achieve a rotational angle larger than that achievable only with legs (e.g., due to to\mbox{rque limita}tion).

\subsection{Proposed Approach and Contribution}

\MF{This work introduces} a compact OCS, based on two flywheels, mounted on the trunk of the lightweight quadruped Solo12 \cite{grimminger2020open}. \MF{The open-source hardware and software are reported in \cite{github_solo}. Figure \ref{fig:solo12flywheels} shows the OCS mounted on the trunk of the robot.} The contributions of the paper are the following:
\begin{itemize}
	\item the design of a novel OCS that enables effective control of the orientation of a legged robot during a jump, while keeping the design simple. 
	In particular, the axes of rotation of the flywheels are set to be incident, enabling \hl{\textit{continuous}} %MDPI: Please confirm if all the italics in the maintext should be retained.
 controllability in both directions of roll and pitch, while keeping the device compact.
	\item simulations  with the quadruped Solo12 were conducted and demonstrated the effectiveness of the proposed approach.
\end{itemize}

\begin{figure}[H]%[t!] 1

	\includegraphics[width=7cm,height=8cm,keepaspectratio]{figures/solo_fw_pic_compressed.png}
	\caption{\small The proposed Orientation Control System for 
		for the $2.5 \ \mathrm{kg}$ quadruped robot Solo12 consists of two $0.1 \ \mathrm{kg}$ flywheels with incident rotation axes. Each wheel is located in a 3D-printed shell and mounted on the trunk of the body.}
	\label{fig:solo12flywheels}
\end{figure}

\subsection{Outline}
The remainder of this paper is organized as follows. In Section \ref{sec:background}, the law of conservation of the total angular momentum is described. In Section \ref{sec:ocs}, the design principles are presented, together with the strategy for simultaneously controlling the robot's roll and pitch orientations. Section \ref{sec:simulations} presents the simulation in different scenarios that demonstrate the capability of our OCS to reject disturbances and to track angular references when there is no contact with the ground, and to dampen base oscillation after touch-down. Conclusions and a possible evolution of the work are reported in Section \ref{sec:conclusions}.

\section{Background}
\label{sec:background}
The starting point for any OCS is the Euler's equation. For any mechanical system, the time derivative of the angular \hl{momentum}, %MDPI: Please confirm if all the bold in equations should be retained.
 $\bm{L}$ computed in respect to a reference point $O$, fixed in an inertial frame, equals the sum of the moments $\bm{M}_i$ applied to the system with respect to the same reference point:
\begin{equation}
\dot{\bm{L}} = \sum_i \bm{M}_i
\end{equation}
\hl{When} %MDPI: Please confirm if the noindent should be retained. The following highlights are the same.
 the result of the external moments applied to the system is zero, and the Euler's equation simplifies to:
 \begin{equation}
\dot{\bm{L}} = 0 \quad \Rightarrow \quad \bm{L}(t) = const,
\label{eq:cons_ang_mom}
\end{equation}
which is known as conservation of angular momentum.
Referring to legged robots, this condition occurs when the system is not in contact with the ground or with other objects, e.g., during a fall or the flight phase of a jump.
In this case, it is possible to change the angular velocity of the base link, thereby changing the joint positions and velocities, as a result of the \textit{non-holonomy} of the total angular momentum \cite{wieber2016modeling}. If the angular momentum of a certain body changes, then one of the others must change to maintain the total sum as constant.
\subsection*{Preliminary Analysis}
Most quadruped robots are designed in such a way that the largest amount of the mass is located in the main body. As a consequence, the contribution of the moments due to the acceleration of other bodies (i.e., leg links) is moderate. \\
As an example, consider the case of Solo12. Even if each limb accounts for about $13.4\%$ of the total mass, changes in the total inertia of the robot come with the motion of only the upper leg (thigh) and the lower leg (calf), since the displacement of distribution of the hip mass with the system's CoM is approximately constant. \MF{In view of this, each limb is responsible for only $7.5\%$ of the total mass in varying the total inertia of the robot}. Flywheels can be used to alleviate the lack of control authority.

\section{Orientation Control System}
\label{sec:ocs}
In this section, a procedure to select the inertia of the flywheels \MF{is presented, together with a strategy} to exploit them so as to simultaneously control the robot's roll and p\mbox{itch orien}tations.

\subsection{Bounds on the Inertia}
Investigation for the selection of the flywheels' inertia can be performed with the Elroy's Beanie model, depicted in Figure \ref{fig:elroy-beanie}.
\begin{figure}[H]%[t!]2

	\includegraphics[width=.7\linewidth]{figures/elroys-beanie.eps}
	\caption{\small Schematic representation of the Elroy's Beanie model used for the preliminary analysis of the pitch motion.}
	\label{fig:elroy-beanie}
\end{figure}

This consists of two rigid bodies connected through their CoMs with a revolute joint. One of the bodies represents the robot in its nominal configuration and the other models the OCS, here represented by a single wheel for the sake of simplicity. The aim was to examine the rotational dynamics of the system as a whole. To simplify the analysis, in the following \MF{only the effects on the pitch angle of the robot was considered}, keeping in mind that the same arguments also apply to the roll. 
Let us identify the moment of inertia of the robot in the nominal configuration as $I_r$ and the moment of inertia of the two flywheels as $I_f$, both referred to the system's CoM.
The angular momentum $L$ of this system can be written as:
\begin{equation}\label{eq:ang_mom_elroy}
L(t) = \left(I_r + I_{f}\right)\dot{\theta}(t) + I_{f}\dot{\gamma}(t)
\end{equation}
where $\dot{\theta}$ and $\dot{\gamma}$ are, respectively, the robot;s pitch rate and the angular speed of the wheel.
\MF{The robot is driven to reach} a desired pitch rate $\dot{\theta}_{des}$ by acting on the wheel speed.
Without loss of generality, \MF{one} can assume the flywheel is stationary with the robot at the instant at which the reorientation maneuver starts ($\dot{\gamma}_0 = 0$), having a system angular momentum of $L_0 = \left(I_r + I_{f}\right)\dot{\theta}_0$.
Under the condition of conservation of the angular momentum, this quantity is constant over time and it is possible to estimate the lower bound for $I_{f}$, given a desired pitch rate and the maximum velocity of the flywheels $\dot{\gamma}_{max}$:
\begin{equation}
I_{f} \geq I_r \dfrac{ \Delta \dot{\theta} }{\Delta \dot{\theta}+\dot{\gamma}_{max}}.
\end{equation}
where $\Delta \dot{\theta} =  \left\lvert \dot{\theta}_0-\dot{\theta}_{des} \right\rvert$ is the base velocity variation.
Figure \ref{fig:inertia_limits} reports the lower bound of $I_{f}$ given the desired base velocity variation and the maximum velocity of the actuator.

\vspace{-9pt}

\begin{figure}[H]%[t]3

	\includegraphics[width=.8\linewidth]{figures/lowerbound2.eps}
	\caption{\small Minimum inertia of the OCS (normalized by the robot's inertia) necessary to achieve a base velocity variation $\Delta \dot{\theta}$, given the actuation bound $\dot \gamma^{max}$.}
	\label{fig:inertia_limits}
\end{figure}

\subsection{Flywheels' Axes of Rotation}
\label{subseq:flywheel_axes}
